---
title: "Verkefni 2"
author: "Jaan Jaerving, email: jaj20@hi.is"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r, setup, results = FALSE, message = FALSE, warning = FALSE}
options(scipen = 999, digits = 12)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(GGally)
library(dplyr)
data(stackloss)
y <- stackloss[, 4]
x1 <- stackloss[, 1]
x2 <- stackloss[, 2]
x3 <- stackloss[, 3]
```

## 1. Start by plotting your data using the `ggpairs()` function from the `GGally` package

```{r}
ggpairs(stackloss)
```

## 2. Fit a multiple linear regression model using the lm() function and store the result in an object called `lm.1`

```{r}
lm.1 <- lm(y ~ x1 + x2 + x3)

s.1 <- summary(lm.1)
s.1
```

## 3. State the estimates for the parameters of the model, the estimate of $\sigma$ and the $R^2$

Intercept of the model is **`r s.1$coefficients[1]`**, slope is **`r s.1$coefficients[2]`**, $\sigma$ is **`r s.1$sigma`** and $R^2$ is **`r s.1$r.squared`**

## 4. State the p-value for the overall F-test $H_0 : \beta_1 = \beta_2 = \beta_3 = 0$

The p-value of the overall F-test is **`r pf(s.1$fstatistic[1], s.1$fstatistic[2], s.1$fstatistic[3], lower.tail = FALSE)`**

## 5. State the p-value for the test $H_0 : \beta_3 = 0$

Estimated p-value for $H_0 : \beta_3 = 0$ is **`r s.1$coefficients[,4][4]`.**

## 6. Estimate the parameters of the model without using the `lm()` function

```{r}
# Add column of 1 values for intercept
X <- as.matrix(cbind(1, x1, x2, x3))
y <- as.matrix(y)

beta_hat <- solve(t(X) %*% X) %*% (t(X) %*% y)
rownames(beta_hat) <- c("Intercept","β1","β2","β3")
beta_hat %>% knitr::kable(caption = "Model parameters without lm()")
```

## 7. Perform an overall F-test using the `anova()` function

```{r}
lm.2 <- lm(y ~ 1)
anova(lm.2, lm.1)
```

## 8. Calculate the value of the test statistic in the test above without using the `anova()` function.

```{r}
rss.1 <- sum(lm.1$res^2)
rss.2 <- sum(lm.2$res^2)
df.1 <- lm.1$df.residual
ms.res <- rss.1 / df.1
ms.diff <- (rss.2 - rss.1) / (lm.2$df.residual - df.1)
f.stat <- ms.diff/ms.res
```

Value of the F statistic is **`r f.stat`**.

## 9. Find a 95% confidence interval for $\beta_3$ using the `confint()` function

```{r}
c1 <- confint(lm.1, "x3")
c1 %>% knitr::kable(caption = "Confint() 95% confidence interval for β3", row.names=FALSE)
```

## 10. Find a 95% confidence interval for $\beta_3$ without using the `confint()` function

```{r}
est <- s.1$coefficients[4,1]
stderr <- s.1$coefficients[4,2]
t.val <- qt(p=.05/2, df=17, lower.tail=FALSE)
data.frame(est - t.val * stderr, est + t.val * stderr) %>%
  knitr::kable(caption = "Manual 95% confidence interval for β3", col.names = c("2.5 %", "97.5 %"),row.names=FALSE)
```