---
title: "Verkefni 2"
author: "Jaan Jaerving"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r, setup, results = FALSE, message = FALSE, warning = FALSE}
options(scipen=999)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(dplyr)
library(GGally)
data(stackloss)
y <- stackloss[, 4]
x1 <- stackloss[, 1]
x2 <- stackloss[, 2]
x3 <- stackloss[, 3]
```

## 1. Start by plotting your data using the `ggpairs()` function from the `GGally` package

```{r}
ggpairs(stackloss)
```

## 2. Fit a multiple linear regression model using the lm() function and store the result in an object called `lm.1`

```{r}
lm.1 <- lm(y ~ x1 + x2 + x3)

s.1 <- summary(lm.1)
s.1
```

## 3. State the estimates for the parameters of the model, the estimate of $\sigma$ and the $R^2$

Intercept of the model is **`r s.1$coefficients[1]`**, slope is **`r s.1$coefficients[2]`**, $\sigma$ is **`r s.1$sigma`** and $R^2$ is **`r s.1$r.squared`**

## 4. State the p-value for the overall F-test $H_0 : \beta_1 = \beta_2 = \beta_3 = 0$

## 5. State the p-value for the test $H_0 : β_3 = 0$

## 6. Estimate the parameters of the model without using the `lm()` function

```{r}
y <- as.matrix(y)
X <- as.matrix(cbind(x1, x2, x3))

betahat <- solve(t(X) %*% X) %*% (t(X) %*% y)
betahat
```

## 7. Perform an overall F-test using the `anova()` function

## 8. Calculate the value of the test statistic in the test above without using the `anova()` function.

## 9. Find a 95% confidence interval for $\beta_3$ using the `confint()` function

## 10. Find a 95% confidence interval for $\beta_3$ without using the `confint()` function