---
title: "Verkefni 3"
author: "Jaan Jaerving (jaj20@hi.is)"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r, setup, results = FALSE, message = FALSE, warning = FALSE}
options(scipen=999)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(ggplot2)
library(GGally)

normalqqplot <- function(x){
  alpha <- 0.05   # set alpha here
  n <- length(x)
  pplot <- (1:n)/(n + 1)
  plow <- qbeta(alpha/2,(1:n),n-(1:n)+1)
  pupp <- qbeta(1-alpha/2,(1:n),n-(1:n)+1)
  qqnorm <- qnorm(pplot)
  qx_low <- qnorm(plow)
  qx_upp <- qnorm(pupp)
  e = (x - mean(x))/sd(x)
  e_sort <- sort(e)
  index_e = order(e)
  figout <- ggplot(data=NULL,aes(x=qqnorm, y=e_sort)) + geom_point() 
  figout <- figout + geom_line(y = qx_low) + geom_line(y = qx_upp)+geom_line(y = qqnorm) 
  figout <- figout + xlab("Standard normal quantiles")+ylab("Standardized residuals")
  figout  <- figout  + geom_text(aes(label = ifelse(e_sort > qx_upp, index_e, "")), hjust = -0.5, vjust=0.4)
  figout  <- figout  + geom_text(aes(label = ifelse(e_sort < qx_low, index_e, "")), hjust = -0.5, vjust=0.4)
  figout
  return(figout)
}

tqqplot <- function(x,df){
  alpha <- 0.05  # set alpha here
  n <- length(x)
  
  pplot <- (1:n)/(n + 1)
  plow <- qbeta(alpha/2,(1:n),n-(1:n)+1)
  pupp <- qbeta(1-alpha/2,(1:n),n-(1:n)+1)
  qqt <- qt(pplot,df)
  qqt_low <- qt(plow,df)
  qqt_upp <- qt(pupp,df)
  e_sort <- sort(x)
  index_e = order(x)
  figout <- ggplot(data=NULL,aes(x=qqt, y=e_sort)) + geom_point() 
  figout <- figout + geom_line(y = qqt_low) + geom_line(y = qqt_upp)+geom_line(y = qqt) 
  figout <- figout + xlab("Standard t quantiles")+ylab("Studentized residuals")
  figout  <- figout  + geom_text(aes(label = ifelse(e_sort > qqt_upp, index_e, "")), hjust = -0.5, vjust=0.4)
  figout  <- figout  + geom_text(aes(label = ifelse(e_sort < qqt_low, index_e, "")), hjust = -0.5, vjust=0.4)
  figout
  return(figout)
}
```

## 0. Read in data, select columns and remove NA values

```{r}
data <- read.table("/home/anjrv/Projects/linear_statistical_models/data/homeprices.csv", 
                   header = TRUE,
                   fill = FALSE,
                   sep = "",
                   dec = ".",
                   na.strings = c("", " ", "NA" , "*"),
                   stringsAsFactors = TRUE)

drop <- c("NE", "CUST", "COR", "TAX")
data <- data[, !(colnames(data) %in% drop)]
data <- na.omit(data)
```

## 1. Use the `ggpairs()` function to get a feel of the data

```{r}
ggpairs(data, lower = list(continuous = wrap("smooth", alpha = 0.3, size=0.2)))
```

In general we can see that `SQFT` has a strong positive effect on `PRICE`. `AGE` on the other hand has a moderate negative effect on `PRICE` values. Properties with more `FEATS` appear to generally trend more expensive but this effect is not sufficient to overcome the other describing variables.

## 2. Fit a linear model with PRICE as response variable and SQFT, AGE and FEATS as predictors.

```{r}
lm.1 <- lm(PRICE ~ SQFT + AGE + FEATS, data)
s.1 <- summary(lm.1)
s.1
```

Looking at the summary of the model there is a question of whether this model should be refitted without `FEATS` included.

## 3. Use the diagnostic tools covered in Chapt. 7 of the Faraway book to detect possible problems with the model. Comment on your findings. Also use the function `tqqplot` in `functions_for_stae312.R` to plot a qq-plot of the studentized residuals (ri) with appropriate bounds.

### Residuals

```{r}
diag <- fortify(lm.1)
diag$jack <- rstudent(lm.1)
p1 <- ggplot(diag, aes(x=seq(1:length(.resid)),y=.resid))+geom_point()
p1 <- p1+geom_hline(yintercept=0, col="red", linetype="dashed")
p1 <- p1+xlab("Index")+ylab("Residuals")
p1 <- p1+geom_text(aes(label=ifelse(abs(.resid)>400,row.names(data),"")),
hjust=-0.1)
p1
```

By looking at residuals from diagnostics we see that we have values at indexes `50`, `89` and `94` of our original data set that might be potential outliers.

### Leverages

```{r}
p <- length(coef(lm.1))
n <- length(fitted(lm.1))
p2 <- ggplot(diag, aes(x=seq(1:length(.hat)),y=.hat))+geom_point()
p2 <- p2+geom_hline(yintercept=2*p/n, col="red", linetype="dashed")
p2 <- p2+xlab("Index")+ylab("Leverages")
p2 <- p2+geom_text(aes(label=ifelse(.hat>2*p/n,row.names(data),"")),
hjust=-0.1)
p2
```

### Studentized residuals

```{r}
p3 <- ggplot(diag, aes(x=seq(1:length(.stdresid)),y=.stdresid))+geom_point()
p3 <- p3+geom_hline(yintercept=0, col="red", linetype="dashed")
p3 <- p3+xlab("Index")+ylab("Studentized residuals")
p3 <- p3+geom_text(aes(label=ifelse(abs(.stdresid)>2,row.names(data),"")),
hjust=-0.1)
p3
```

Moving to studentized residuals we see the same three values `50`, `89` and `94` deviate substantially from the group. `89` specifically appears to be a strong candidate for an outlier value.

### Jacknife residuals

```{r}
p4 <- ggplot(diag, aes(x=seq(1:length(jack)),y=jack))+geom_point()
p4 <- p4+geom_hline(yintercept=0, col="red", linetype="dashed")
p4 <- p4+xlab("Index")+ylab("Jackknife residuals")
p4 <- p4+geom_text(aes(label=ifelse(abs(jack)>2,row.names(data),"")),
hjust=-0.1)
p4
```

Jacknife largely confirms what we have seen from residuals before.

```{r}
p5 <- ggplot(diag, aes(x=seq(1:length(.cooksd)),y=.cooksd))+geom_point()
p5 <- p5+xlab("Index")+ylab("Cooks distance")
p5 <- p5+geom_text(aes(label=ifelse(.cooksd>0.15,row.names(data),"")),
hjust=-0.1)
p5
```

Cooks distance shows us observations that are influential on the model. Here we see the three values values that we have noted previously but also the value at index `43` of the original data. Here we see that of the three potential outlier values the value at `94` would have the largest effect on the model.

```{r}
lm.2 <- lm(PRICE ~ SQFT + AGE + FEATS, diag, subset = (.cooksd < max(.cooksd)))
s.2 <- summary(lm.2)
s.2
```

Removing `94` with the highest Cook value changes our intercept rather dramatically and makes it even more likely that we should perhaps move to a model without the `FEATS` column.

### More residual plots

```{r}
p6<-ggplot(diag, aes(x=.fitted,y=.resid))+geom_point()
p6<-p6+stat_smooth(method="loess",se=F)+
geom_hline(yintercept=0, col="red", linetype="dashed")
p6<-p6+xlab("Fitted")+ylab("Residuals")
p6
```

The data is mostly scattered around the vertical line although there is a question of whether there is non-constant variance when looking at how the data scatters as price increases and in that case a transform could be applied to the data.

### QQ-plots

```{r}
p10 <- ggplot(diag, aes(sample = .stdresid)) + stat_qq()
p10 <- p10 + geom_abline(slope=1)
p10 <- p10 + ylab("Studentized residuals")
p10
```

Most values fall around the middle line of the graph but there appears to be a curve off at the extremities. This indicates that the outlier values we have been seeing in our other plots are more extreme than would be expected if they came from a normal distribution.

```{r}
par(mfrow=c(2,2))
plot(lm.1)
par(mfrow=c(1,1))
```

### Tqqplot

```{r}
tqqplot(diag$.stdresid, 15)
```

