---
title: "Verkefni 3"
author: "Jaan Jaerving (jaj20@hi.is)"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r, setup, results = FALSE, message = FALSE, warning = FALSE}
options(scipen=999)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)

library(GGally)

data <- read.table("/home/anjrv/Projects/linear_statistical_models/data/homeprices.csv", 
                   header = TRUE,
                   fill = FALSE,
                   sep = "",
                   dec = ".",
                   na.strings = c("", " ", "NA" , "*"),
                   stringsAsFactors = TRUE)
```

## 1. Use the `ggpairs()` function to get a feel of the data

```{r}
ggpairs(data)
```

## 2. Fit a linear model with PRICE as response variable and SQFT, AGE and FEATS as predictors.

```{r}
lm.1 <- lm(data$PRICE ~ data$SQFT + data$AGE + data$FEATS)
s.1 <- summary(lm.1)
s.1
```

## 3. Use the diagnostic tools covered in Chapt. 7 of the Faraway book to detect possible problems with the model. Comment on your findings. Also use the function `tqqplot` in `functions_for_stae312.R` to plot a qq-plot of the studentized residuals (ri) with appropriate bounds.

```{r}

```
