---
title: "Verkefni 1"
author: "Jaan Jaerving"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r, setup, results = FALSE, message = FALSE, warning = FALSE}
options(scipen=999)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

## 1. Start by plotting your data using the `plot()` function

```{r}
data <- read.table("/home/anjrv/Projects/linear_statistical_models/data/bjor.csv", 
                   header = TRUE,
                   fill = FALSE,
                   sep = ";",
                   dec = ".",
                   na.strings = c("", " ", "NA"),
                   stringsAsFactors = TRUE)

plot(data)
```

## 2. Fit a linear regression model using the `lm()` function and store the result in an object called `fit.lm`. You can see the results using the `summary()` function.

```{r}
fit.lm <- lm(amagn~fjbjor, data)
summary(fit.lm)
```

## 3. Give estimates of the parameters of the model along with the value of the coefficient of determination.



## 4. Use the `predict()` function to predict the alcohol consumption after drinking 4.5 beers

```{r}
var_beer <- data.frame(fjbjor=c(4.5))
predict(fit.lm, var_beer)
```

## 5. Give a 95% prediction interval at x~h~ = 4.5

```{r}
predict(fit.lm, var_beer, interval = 'confidence')
```
